---
title: "DATA607 Week 5 Assignment"
author: "ODONNELL"
date: "February 26, 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

First, import needed libraries
```{r}
library(readr)
library(tidyr)
library(dplyr)
```

Now, load the data just as it was posted
And, view it.
```{r}
flights <- read_csv("week5_assignment_data.csv")
show(flights)
```

OK, the data's pretty messy so:
1. rename column 1
2. rename column 2
3. delete empty row
4. fill in empty cells
```{r}
names(flights)[1]<-"Airline"

names(flights)[2]<-"Arrival_Status"

flights <- flights[-c(3),]

flights[2, "Airline"] <- "ALASKA"
flights[4, "Airline"] <- "AM WEST"
```

Now, start tidying the data with tidyr and dplyr
First, gather the city columns together and arrange them
```{r}
flights <- gather(flights, "Arrival_City", "Freq", 3:7)
flights <- arrange(flights, Airline)

show(flights)
```

Quickly, lets make sure all the columns are appropriate data types
```{r}
str(flights)
```



--------------------------------------------------------------------
Now, lets use tidyr and dyplr to do some analysis:
First, lets compare the total number of flights between the airlines
```{r}
flights %>%
  group_by(Airline) %>%
  summarise(total_flights = sum(Freq))
```


Next, lets compare percentage of delayed flights between ALASKA and AM WEST
```{r}
flights %>%
  filter(Airline == 'ALASKA') %>%
  group_by(Arrival_Status) %>%
  summarise(total_flights = sum(Freq), percentage = sum(Freq)/sum(subset(flights, Airline == 'ALASKA')$Freq))

flights %>%
  filter(Airline == 'AM WEST') %>%
  group_by(Arrival_Status) %>%
  summarise(total_flights = sum(Freq), percentage = sum(Freq)/sum(subset(flights, Airline == 'AM WEST')$Freq))
```

Next, lets compare percentage of delayed flights between ALASKA and AM WEST
by arrival city. There may be variance among cities.
```{r}
cities <- spread(flights, "Arrival_City", "Freq") %>%
  filter(Airline == "ALASKA")
cities <- mutate(cities, Seattle = Seattle/sum(subset(flights, Airline == 'ALASKA' & Arrival_City == "Seattle")$Freq))
cities <- mutate(cities, `Los Angeles` = `Los Angeles`/sum(subset(flights, Airline == 'ALASKA' & Arrival_City == "Los Angeles")$Freq))
cities <- mutate(cities, Phoenix = Phoenix/sum(subset(flights, Airline == 'ALASKA' & Arrival_City == "Phoenix")$Freq))
cities <- mutate(cities, `San Diego` = `San Diego`/sum(subset(flights, Airline == 'ALASKA' & Arrival_City == "San Diego")$Freq))
cities <- mutate(cities, `San Francisco` = `San Francisco`/sum(subset(flights, Airline == 'ALASKA' & Arrival_City == "San Francisco")$Freq))

cities_amwest <- spread(flights, "Arrival_City", "Freq") %>%
  filter(Airline == "AM WEST")
cities_amwest <- mutate(cities_amwest, Seattle = Seattle/sum(subset(flights, Airline == 'AM WEST' & Arrival_City == "Seattle")$Freq))
cities_amwest <- mutate(cities_amwest, `Los Angeles` = `Los Angeles`/sum(subset(flights, Airline == 'AM WEST' & Arrival_City == "Los Angeles")$Freq))
cities_amwest <- mutate(cities_amwest, Phoenix = Phoenix/sum(subset(flights, Airline == 'AM WEST' & Arrival_City == "Phoenix")$Freq))
cities_amwest <- mutate(cities_amwest, `San Diego` = `San Diego`/sum(subset(flights, Airline == 'AM WEST' & Arrival_City == "San Diego")$Freq))
cities_amwest <- mutate(cities_amwest, `San Francisco` = `San Francisco`/sum(subset(flights, Airline == 'AM WEST' & Arrival_City == "San Francisco")$Freq))

#cities
#cities_amwest

cities <- bind_rows(cities, cities_amwest)
cities <- cities[-c(2,4),]

cities
```

Now, lets compare the amount of flights to each
Arrival City between the Airlines
```{r}
flights %>%
  filter(Airline == 'ALASKA') %>%
  group_by(Arrival_City) %>%
  summarise(total_flights = sum(Freq), percentage = sum(Freq)/sum(subset(flights, Airline == 'ALASKA')$Freq))

flights %>%
  filter(Airline == 'AM WEST') %>%
  group_by(Arrival_City) %>%
  summarise(total_flights = sum(Freq), percentage = sum(Freq)/sum(subset(flights, Airline == 'AM WEST')$Freq))
```

Conclusions:
1. AM WEST has 3432 more flights in the dataset than ALASKA
2. The airline ALASKA, has 2.79% more delayed flights overall than AM WEST
3. Among cities, the airlines have very different percentages of delayed flights,
  for example, AM WEST has 9.5% more delayed flights into San Francisco than ALASKA
4. In this dataset, the airlines have a different distribution of flights into
  each city. or example, ALAKSA mostly flys into San Francisco while AM WEST
  mostly flys into Pheonix.
  
Note: all conclusions made based on the 4 R chunks above!